---
title: "Weather events with the greatest health and economic impacts across the
United States, from 1995-2011"
output: html_document
---

#Synopsis

This report aims to answer the question, "which types of storms have the 
greatest health and economic impact?"

Data was gathered from the U.S. National Oceanic and Atmospheric Administration 
(NOAA) storm database. Specifically, data indicating the  type, date, numbers of 
fatalities and injuries, and extent of property and crop  damage for a given 
storm was used to answer the question.

Analysis of the data indicates that storms characterized as Hurricanes/Typhoons
have the most negative economic impact, principally in the form of property 
damage, and those characterized as Floods have the most negative impact on the
health of a population, principally through fatalities. Closer investigation
reveals that 72% of total economic damage is caused by one type of storm - 
Hurricane/Typhoon, and 80% of total health damage is caused by three types of 
storms - Floods, Tornadoes, and Hurricane/Typhoons. It is interesting to note 
that the most economically damaging storms are not necessarily the most life
-threatening storms. 

Finally, deeper examination of the Hurricane/Typhoon category shows that the
economic damage caused by Hurricane/Typhoons from 1995-2011 was most likely
skewed by one hurricane which ended on August 29, 2005 - which is also the end
date of Hurricane Katrina.

The code and figures providing evidence and more detailed discussion of the 
possible conclusions suggested here are found below.

---

##Results

The data was analyzed for overall damage to health - by fatality or injury - and
overall damage to economy - through property or crop damage. A panel of 
histograms was created (using ggplot2) to look at the sum of health vs. economic
damages caused by a given weather event.

NOTE: The below code allows for manipulation of the x-axis, whereby the histogram
bars are ordered in decreasing frequency.

```{r test, echo=TRUE, cache=TRUE}
#Use for generating factor levels when ordering weather events in qplot

healthSums <- stormdata[DamageType == "Health"][, sum(Damage), by = Event]
economicSums <- stormdata[DamageType == "Economic"][, sum(Damage), by = Event]
sums <- merge(healthSums, economicSums, by = "Event")
setnames(sums, old = c("V1.x", "V1.y"), new = c("healthSums", "economicSums"))
stormdata <- merge(stormdata, sums, by = "Event")

count <- data.frame(table(stormdata$Event))
setnames(count, old = "Var1", new = "Event")
stormdata <- merge(x = stormdata, y = count, by = "Event")
```

```{r results2, echo=TRUE, cache=TRUE, eval=TRUE}
healthFactLvls <- sums[order(healthSums, decreasing = TRUE)]$Event
economicFactLvls <- sums[order(economicSums, decreasing = TRUE)]$Event

ggplot(data = stormdata, aes(fill = EventType, alpha = DamageType2, 
                             weight = Damage, color = EventType)) +
        geom_bar(data = stormdata[DamageType == "Health"], 
                 aes(x = factor(Event, levels = healthFactLvls))) +
        geom_bar(data = stormdata[DamageType == "Economic"], 
                 aes(x = factor(Event, levels = economicFactLvls))) +
        facet_wrap(~ DamageType, scales = "free") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
              panel.background = element_rect("white")) +
        scale_fill_hue(l = 45) +
        labs(x = "Storm Type", 
             y = "Total Damage",
             title = "Total Damage per Storm Type, by type of consequence")
```

In the above plots, total damage, expressed in dollars for the economic data
and counts for the health data, is graphed against weather event. Each bar is
stacked to show the sum of crop and property damage in the case of the economic
data and fatalities and injuries in the case of the health data. Shading has
been applied to demarcate these stacked values. Further, colors have been
applied to the umbrella categories of weather events: Flood, Heat, Rain, Wind, 
and Winter.

As the histograms clearly indicate, weather events falling under the category
"Hurricane/Typhoon" have the most harmful economic consequences, primarily in 
the form of property damage. Weather events falling under the category 
"Flood" have the most harmful health consequences, primarily in the form of
fatalities.

It is interesting to note that the most harmful weather events with respect to
human lives are not the most harmful weather events with respect to property. 
Further, the data reflecting economic consequences is heavily skewed right, 
begging the question as to what has caused the skew. Is the **number** of 
hurricanes in the US from 1995-2011 dramatically greater than any other weather 
event, or is the **impact** of a few hurricanes large enough to offset that of 
other events?

To address this question, the data was graphed in a different manner, to reflect 
the frequency of each weather event as well.

```{r results3, echo=TRUE, cache=TRUE, eval=TRUE}
ggplot(stormdata) +
        geom_point(data = stormdata[DamageType == "Health"],
                   aes(x = Event, y = healthSums, color = EventType, 
                       size = Freq)) +
        geom_point(data = stormdata[DamageType == "Economic"],
                   aes(x = Event, y = economicSums, color = EventType, 
                       size = Freq)) +
        facet_wrap(~ DamageType, scales = "free") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
        scale_fill_hue(l = 45) +
        labs(x = "Storm Type",
             y = "Total Damage",
             title = "Total Damage per Storm Type, by type of consequence")
```

In the above plots, total damages are again plotted against weather event. In 
this graph, color indicates the umbrella categories of weather events, and the 
size of each point indicates the frequency of a given weather event.

The graphs show that Tornados have occured with the greatest frequency across
the US. However, the greatest amount of damage is not caused by tornadoes 
either in economic or in health terms. In fact, the event type with the 
greatest economic damage - Hurricane/Typhoon - has a very low frequency. 
Similarly the event type with the greatest health damage - Flood - has a low
frequency as well. Further, the Hurricane/Typhoon category accounts for 72% of 
total economic damage and the Flood, Tornado, and Hurricane/Typhoon categories 
account for 80% of total health damage. Thus, it is clear that frequency of 
weather events does not necessarily translate to greatest damage.

According to closer examination of the hurricane/typhoon data, there have 
been 12 events in this category from 1995-2011. The event which caused the
greatest property damage ended on 08-29-2005 - the date of Hurricane Katrina's
second landfall, just before it was downgraded to a tropical storm. This is 
likely the event which is skewing the data.  

```{r results4, echo=TRUE, cache=TRUE, eval=TRUE}
stormdata[Event == "HURRICANE/TYPHOON"][which(Damage == max(Damage))]

econTot <- sum(stormdata[DamageType == "Economic"]$Damage)
healthTot <- sum(stormdata[DamageType == "Health"]$Damage)

econ <- function(x){
        (x/econTot)*100
}

health <- function(x){
        (x/healthTot)*100
}

unique(stormdata[, econ(economicSums), by = Event][order(-V1)])
unique(stormdata[, health(healthSums), by = Event][order(-V1)])
```

---

##Data Processing

###Loading data into R

The data was downloaded and read into R using the code below. 

```{r setup, echo=TRUE, cache=TRUE}
library(data.table)
library(lattice)
library(ggplot2)
options(datatable.print.topn = 5)

setwd("F:/Data_science_coursera/Reproducible_Research/Storm_Analysis")
```

```{r loading, echo=TRUE, cache=TRUE}
download.file(url = "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
              destfile = "storm.csv")
```

###Pre-processing

The data was stored as a data table using the data.table package, to ease 
manipulation. After reading in the data, some exploratory data analysis was 
conducted to see the dimensions of the data table, assess for missing values, 
and understand more about the variables.

```{r pre-processing, echo=TRUE, cache=TRUE}
data <- read.csv("storm.csv")
```

```{r pre-processing2, echo=TRUE, cache=TRUE}
dim(data)
names(data)
```

The raw data requires a significant amount of cleaning for the purpose of this 
analysis, which focuses on the population health and economic consequences of 
storm events across the US. 

The code description is contained in the descriptive document which can be 
downlaoded here: 
https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

This information was used to select the following variables as relevant to 
retain in the dataset:

* (2) BGN_DATE: Date and time the weather event began
* (8) EVTYPE: Type of storm
* (12) END_DATE: Date and time the weather event ended
* (23) FATALITIES: Number of fatalities resulting from storm
* (24) INJURIES: Number of injuries resulting from storm
* (25) PROPDMG: Number value of amount of property damage
* (26) PROPDMGEXP: Coefficient correlating to 10^3 (K), 10^6 (M), or 10^9 (B) 
                        times the number value for property damage
* (27) CROPDMG: Number value of amount of crop damage
* (28) CROPDMGEXP: Coefficient correlating to 10^3 (K), 10^6 (M), or 10^9 (B) 
                        times the number value for crop damage
                        
The information on dates was retained primarily for exploratory reasons, as the
codebook indicated that information has been generally missing in early years.

```{r pre-processing3, echo=TRUE, cache=TRUE}
stormdata <- data.table(data[c(2, 8, 12, 23:28)])
```

The data was examined for NA values, which were not found.

```{r pre-processing4, echo=TRUE, cache=TRUE}
tables()
length(which(is.na(stormdata) == TRUE))
stormdata
```

Transformation on the data was performed with respect to three variables 
(1) year, (2) weather event, and (3) damages. As the below code will show, these
variables were both cleaned for invalid/unnecessary observations and transformed 
for better analysis. 

(1) Only data for years with the most reliable and complete information 
(according to the codebook and the below exploratory data analysis) was kept.
(2) Only weather events which resulted in damages, either in the form of 
fatality, injury, property damage, or crop damage were kept. The weather events 
were also grouped under the umbrella categories, Heat, Flood, Rain, Wind, and 
Winter.
(3) Only valid entries for the property and crop damage factor of ten (K M or B, 
according to the codebook) were retained. All damages (fatality, injury, 
property and crop) were also combined into one variable with four levels.

NOTE: As the below code will indicate, the damages variables were cleaned as a 
consequence of cleaning the events variable. For this reason, the events 
variable was cleaned first.

###Transforming the data - *years*

The data was transformed so that the date variables could be examined for the
most useful years to study. The times have been removed as that level of 
specificity is not necessary for the analysis in this report.

```{r transformation, echo=TRUE, cache=TRUE}
stormdata$BGN_DATE <- as.Date(stormdata$BGN_DATE, format = "%m/%d/%Y")
stormdata$END_DATE <- as.Date(stormdata$END_DATE, format = "%m/%d/%Y")

beginYr <- as.numeric(format(stormdata$BGN_DATE, "%Y"))
endYr <- as.numeric(format(stormdata$END_DATE, "%Y"))

bYears <- data.table("Date" = beginYr, 
                      "BGN.END" = rep("BEG", length(beginYr)))
eYears <- data.table("Date" = endYr, 
                    "BGN.END" = rep("END", length(endYr)))
years <- rbind(bYears, eYears)
years
histogram(~Date|BGN.END, data = years, type = "density", 
          scales = list(x = list(at = seq(1950, 2010, by = 5), rot = 45)),
          main = "Density of datapoints per year", xlab = "Year")
```

The above histograms show the density of recorded datapoints per year.

The figure clearly show that in the years prior to 1995, beginning date entries 
were more sparse, and end date entries were mostly non-existent. Thus, the most 
complete information is available for years after 1995. The data was then 
subsetted for 1995 onwards. The end date was  arbitrarily selected as the 
criterion for subsetting and for marking the event. As such, a given weather 
event in this analsyis is measured by the date it subsided.

```{r transformation2, echo=TRUE, cache=TRUE}
stormdata <- stormdata[which(as.numeric(format(END_DATE, "%Y")) >= 1995)]
stormdata$BGN_DATE <- NULL
setcolorder(stormdata, c("END_DATE", "EVTYPE", "FATALITIES", 
                         "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", 
                         "CROPDMGEXP"))
setnames(stormdata, old = names(stormdata), new = c("Date", "Event",
                                                    "Fatalities", "Injuries",
                                                    "Propdamage", "PROPDMGEXP",
                                                    "Cropdamage", "CROPDMGEXP"))
rm(bYears)
rm(eYears)
rm(years)
tables()
```

###Transforming the data - *events*
The Event variable was examined for unnecessary information and for overall
categories. 

```{r transformation3, echo=TRUE, cache=TRUE}
length(unique(stormdata$Event))
head(unique(stormdata$Event))
tail(unique(stormdata$Event))
```

As this analysis focuses on the health and economic consequences of
a given weather event, those observations referring to summaries were removed
from the dataset, as were those resulting in none of the following: fatalities, 
injuries, property damage, or crop damage.

```{r transformation4, echo=TRUE, cache=TRUE}
stormdata <- stormdata[!grep("Summary", Event)]

setkey(stormdata, Event)

stormdata <- stormdata[which(Fatalities > 0 & Injuries > 0 & Propdamage > 0 & 
                                     Cropdamage > 0)]
tables()

unique(stormdata$Event)
```

The resulting list of 86 events were transformed to lump events with very 
similar notation together (i.e. TSTM WIND was grouped with Thunderstorm wind, 
Hurricane/Typhoon was grouped with Hurricane, and Excessive heat was grouped 
with Heat.) A factor variable was also added tothe data, providing an umbrella 
category for the weather event. The following umbrella categories were applied, 
by studying the remaining event types in the dataset, the original guidelines 
provided to data collectors in the codebook, and research on the National 
Oceanic and Atmospheric Administration site (http://www.NOAA.gov):

* Heat events (HEAT, HEAT WAVE DROUGHT, WILDFIRE)
* Flood events (FLOOD, FLASH FLOOD, TSUNAMI)
* Rain events (HURRICANE/TYPHOON, TROPICAL STORM)
* Wind events (HIGH WIND, THUNDERSTORM WIND, TORNADO, WINTER STORM HIGH WINDS)
* Winter events (HEAVY SNOW, WINTER STORM)

```{r transformation5, echo=TRUE, cache=TRUE}
stormdata[Event == "TSTM WIND"]$Event <- "THUNDERSTORM WIND"
stormdata[Event == "HURRICANE"]$Event <- "HURRICANE/TYPHOON"
stormdata[Event == "EXCESSIVE HEAT"]$Event <- "HEAT"
unique(stormdata$Event)

stormdata$EventType <- ""

Heat <- c("HEAT", "HEAT WAVE DROUGHT", "WILDFIRE")
Flood <- c("FLOOD", "FLASH FLOOD", "TSUNAMI")
Rain <- c("HURRICANE/TYPHOON", "TROPICAL STORM")
Wind <- c("HIGH WIND", "THUNDERSTORM WIND", "TORNADO", "WINTER STORM HIGH WINDS")
Winter <- c("HEAVY SNOW", "WINTER STORM")

stormdata[Event%in%Heat]$EventType <- "Heat"
stormdata[Event%in%Flood]$EventType <- "Flood"
stormdata[Event%in%Rain]$EventType <- "Rain"
stormdata[Event%in%Wind]$EventType <- "Wind"
stormdata[Event%in%Winter]$EventType <- "Winter"

stormdata$EventType <- as.factor(stormdata$EventType)
head(stormdata)
tail(stormdata)
```

###Transforming the data - *damages*

The data on damages was transformed in two ways: (1) to correct the economic 
damages (property and crop damage) for the factor of ten provided in the 
variables PROPDMGEXP and CROPDMGEXP, and (2) to combine the damage variables
as one variable with four levels: Fatality, Injury, Property and Crop.
These transformations were performed to allow for more elegant graphical 
analysis.

(1) Correcting for factor of 10 in economic damages

```{r transformation6, echo=TRUE, cache=TRUE}
unique(stormdata$PROPDMGEXP)
unique(stormdata$CROPDMGEXP)

stormdata$PEXP <- 0
stormdata$CEXP <- 0

stormdata[PROPDMGEXP == "K"]$PEXP <- 10^3
stormdata[PROPDMGEXP == "M"]$PEXP <- 10^6
stormdata[PROPDMGEXP == "B"]$PEXP <- 10^9

stormdata[CROPDMGEXP == "K"]$CEXP <- 10^3
stormdata[CROPDMGEXP == "M"]$CEXP <- 10^6
stormdata[CROPDMGEXP == "B"]$CEXP <- 10^9

stormdata$Propdamage <- stormdata$Propdamage*stormdata$PEXP
stormdata$Cropdamage <- stormdata$Cropdamage*stormdata$CEXP

stormdata$PROPDMGEXP <- NULL
stormdata$CROPDMGEXP <- NULL
stormdata$PEXP <- NULL
stormdata$CEXP <- NULL

damages <- rbind(data.table("Damage" = stormdata$Fatalities,
                            "DamageType" = "Health",
                            "DamageType2" = "Fatality"), 
                 data.table("Damage" = stormdata$Injuries,
                            "DamageType" = "Health",
                            "DamageType2" = "Injury"),
                 data.table("Damage" = stormdata$Propdamage,
                            "DamageType" = "Economic",
                            "DamageType2" = "Property"),
                 data.table("Damage" = stormdata$Cropdamage,
                            "DamageType" = "Economic",
                            "DamageType2" = "Crop"))

stormdata <- cbind(stormdata, damages)

stormdata$Fatalities <- NULL
stormdata$Injuries <- NULL
stormdata$Propdamage <- NULL
stormdata$Cropdamage <- NULL

stormdata$DamageType <- as.factor(stormdata$DamageType)
stormdata$DamageType2 <- as.factor(stormdata$DamageType2)
```

The final dataset is completely clean.

```{r transformation7, echo=TRUE, cache=TRUE}
stormdata
```